---
title: "Beta-Carotene Mixed-Effects Analysis (NIMBLE)"
author: "Miheer Dewaskar (adapted from Prof. Fletcher Christensen)"
---

```{r setup}
library(nimble)
library(coda)
```

## Data Import

Update `bc_datafile` to match your local path.

```{r data}
bc_datafile <- "bcarotene.csv"   # <-- update path as needed

bc_data <- read.csv(bc_datafile)
```

## Data Cleaning

Note first that different patients in the study have data for a different number of months.

```{r}
library(dplyr)

bc_data |>
  group_by(ptid) |> summarize(obs.count = n())
```

Some patients in fact even have their month number repeating (whenever num_obs > num_months).

```{r}
bc_data |> group_by(ptid) |> summarize(num.obs = n(), num.months = n_distinct(month))
```

Let's see which patients received what doeses

```{r}
bc_data |> group_by(ptid) |> summarize(doses = unique(dose))
```

Note that months 0-3 are supposed to be placebo for everybody but their dosage does not say this.

```{r}
bc_data |> 
  filter(month %in% 0:3) |> 
    group_by(ptid) |> 
      summarize(zero_count = sum(dose==0))
```

Let's plot bcarot levels for patient versus time, along with their average levels. We should expect to see a "jump" around month 4.

```{r}
library(ggplot2)

for(ptid.s in unique(bc_data$ptid)) {
  dose.s <- unique(bc_data[bc_data$ptid == ptid.s, "dose"])
  age.s <- unique(bc_data[bc_data$ptid == ptid.s, "age"])
  male.s <- unique(bc_data[bc_data$ptid == ptid.s, "male"])

  bc_data |> 
    filter(ptid == ptid.s) |>
      ggplot(aes(y=bcarot, x=month)) + geom_point() + 
        # show patient information in the title
        ggtitle(sprintf("Patient %d: Dose = %d, Age = %d, Male = %d", ptid.s, dose.s, age.s, male.s)) -> g
  print(g)
}
```

We can see that some months repeat and have abnormally high bcarot value (compare these to AUC). You can remove such values, but must clearly document this.

## Model fitting

```{r}
ptid   <- bc_data$ptid
month  <- bc_data$month
bcarot <- bc_data$bcarot
vite   <- bc_data$vite
dose   <- bc_data$dose
age    <- bc_data$age
male   <- bc_data$male
bmi    <- bc_data$bmi
chol   <- bc_data$chol
cauc   <- bc_data$cauc
vauc   <- bc_data$vauc
```

## Derived Variables and Design Matrix

This is **a demo model** that shows how to use NIMBLE, not necessarily **the model that I recommend using**. Watch Prof. Fletcher Christensen's second and third videos to understand what this is doing.

```{r design}
n          <- nrow(bc_data)
n_patients <- length(unique(ptid))

intercept <- rep(1, n)
tx        <- intercept - (month < 4)   # 0 before month 4, 1 from month 4 onward

x            <- cbind(intercept, tx * dose, age)
n_covariates <- ncol(x)

tau_b <- 0.00001   # flat prior precision for fixed effects
```

## MCMC Parameters

These are just placeholds, and might need to change these parameters based on MCMC diagnostics.

```{r mcmc-params}
chains     <- 1
burn_in    <- 1000
iterations <- 100000
thin       <- 1
```

## NIMBLE Model

This is a basic demo of a `random effects` model with `x` denoting covariate matrix (see discussion above where this was defined).

```{r model}
bc_code <- nimbleCode({

  ## Likelihood
  for (i in 1:n) {
    bcarot[i] ~ dnorm(mu[i], tau_bc)
    mu[i]     <- inprod(x[i, 1:n_covariates], beta[1:n_covariates]) + gamma[ptid[i]]
  }

  ## Patient-level random effects
  for (j in 1:n_patients) {
    gamma[j] ~ dnorm(0, tau_g)
  }

  ## Fixed-effects priors
  for (k in 1:n_covariates) {
    beta[k] ~ dnorm(0, tau_b)
  }

  ## Precision priors
  tau_bc ~ dgamma(0.001, 0.001)
  tau_g  ~ dgamma(0.001, 0.001)

  ## Derived standard deviations
  sigma_bc <- pow(tau_bc, -0.5)
  sigma_g  <- pow(tau_g,  -0.5)

})
```

## Constants, Data, and Inits

NIMBLE distinguishes *constants* (values that index loops or define fixed structure)
from *data* (observed stochastic nodes). Mixing them causes silent errors, so they
are kept separate here.

```{r data-lists}
## Constants: anything that appears as a loop bound or fixed hyperparameter
bc_constants <- list(
  n            = n,
  n_patients   = n_patients,
  n_covariates = n_covariates,
  tau_b        = tau_b,
  ptid         = ptid,    # used as an index into gamma[]
  x            = x
)

## Data: observed stochastic nodes
bc_data_list <- list(
  bcarot = bcarot
)

## Initial values
bc_inits <- list(
  beta   = rnorm(n_covariates, 0, 1),
  gamma  = rnorm(n_patients,   0, 1),
  tau_bc = runif(1, 0, 2),
  tau_g  = runif(1, 0, 2)
)
```

## Run MCMC

```{r build, results='hide'}
## If you don't need custom samplers, nimbleMCMC() is a shorter wrapper
## around model building, MCMC configuration, compilation, and execution.
set.seed(42)

bc_samples <- nimbleMCMC(
  code      = bc_code,
  constants = bc_constants,
  data      = bc_data_list,
  inits     = bc_inits,
  monitors  = c("beta", "gamma", "sigma_bc", "sigma_g"),
  niter     = iterations,
  nburnin   = burn_in,
  nchains   = chains,
  thin      = thin,
  setSeed   = FALSE,   # using set.seed(42) above
  samplesAsCodaMCMC = TRUE
)
```


## Extract Posterior Samples

```{r extract}
## runMCMC with nchains=1 returns a single mcmc object;
## with nchains>1 it returns an mcmc.list. Handle both.
if (chains == 1) {
  bc_iterates <- as.data.frame(as.matrix(bc_samples))
} else {
  bc_iterates <- as.data.frame(as.matrix(bc_samples))  # combines all chains
}

## Column names follow NIMBLE's convention: "beta[1]", "gamma[1]", etc.
beta_cols  <- paste0("beta[",  1:n_covariates, "]")
gamma_cols <- paste0("gamma[", 1:n_patients,   "]")
sigma_cols <- c("sigma_bc", "sigma_g")

posterior_betas  <- bc_iterates[, beta_cols,  drop = FALSE]
posterior_gammas <- bc_iterates[, gamma_cols, drop = FALSE]
posterior_sigmas <- bc_iterates[, sigma_cols, drop = FALSE]
```

## Posterior Summaries

### Fixed Effects

```{r summary-beta}
summary_betas <- t(apply(posterior_betas, 2, function(x)
  c(Mean   = mean(x),
    SD     = sd(x),
    `2.5%` = unname(quantile(x, 0.025)),
    `97.5%`= unname(quantile(x, 0.975)))))

knitr::kable(summary_betas, digits = 4,
             caption = "Posterior summary: fixed effects (beta)")
```

### Variance Components

```{r summary-sigma}
summary_sigmas <- t(apply(posterior_sigmas, 2, function(x)
  c(Mean    = mean(x),
    SD      = sd(x),
    `2.5%`  = unname(quantile(x, 0.025)),
    `97.5%` = unname(quantile(x, 0.975)))))

knitr::kable(summary_sigmas, digits = 4,
             caption = "Posterior summary: residual and random-effect SDs")
```

### Random Effects (first 10 patients)

```{r summary-gamma}
summary_gammas <- t(apply(posterior_gammas[, 1:min(10, n_patients), drop=FALSE], 2, function(x)
  c(Mean    = mean(x),
    SD      = sd(x),
    `2.5%`  = unname(quantile(x, 0.025)),
    `97.5%` = unname(quantile(x, 0.975)))))

knitr::kable(summary_gammas, digits = 4,
             caption = "Posterior summary: patient random effects, first 10 patients")
```

## MCMC Diagnostics

```{r diagnostics, fig.height=5}
## Trace and density plots for key parameters
plot(bc_samples[, beta_cols,  drop = FALSE], ask = FALSE)
plot(bc_samples[, sigma_cols, drop = FALSE], ask = FALSE)
```

```{r ess}
## Effective sample size
ess <- effectiveSize(bc_samples[, c(beta_cols, sigma_cols), drop = FALSE])
knitr::kable(data.frame(Parameter = names(ess), ESS = round(ess)),
             row.names = FALSE,
             caption = "Effective sample size for fixed effects and variance components")
```
